---
title: Lab 5
author: Hunter Rose, Tanner Boyle
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
```

```{r}
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv")

person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv")
```

```{r}
acc %>%
  select(DAY_WEEK) %>%
  table()
```
As we can see, day 7 is the day with the most crashes. According to the FARS manual, day 7 is Saturday.
```{r}
acc %>%
  select(HOUR) %>%
  table()
```
The FARS manual declares the time variable in military time, and the hour with the most crashes is 18, or 6 o'clock. This makes sense because it is within rush hour which is usually between 4-8pm.
```{r}
acc %>%
  select(DRUNK_DR) %>%
  table()
```
This table shows us that the majority of crashes do not have drunk drivers. The crashes that do, tend to only have one, with only a few cases having more than 1. The total number of crashes with at least one drunk driver is 8769.
```{r}
colnames(person)
colnames(acc)
```